---
title: "Fly Mating: evolution rate figs"
author: "Aram Avila-Herrera"
date: "`r date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 4
    highlight: pygments
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = '#>')

# from SO user DeanK: https://stackoverflow.com/a/18672268
tn = local({
  i = 0
  function(x) {
    i <<- i + 1
    paste('\n\n:Table ', i, ': ', x, sep = '')
    # The : before Table tells pandoc to wrap your caption in <caption></caption>
  }
})
knitr::knit_hooks$set(tab_cap = function(before, options, envir) {
  if(!before)
    tn(options$tab_cap)
})
default_output_hook = knitr::knit_hooks$get("output")
knitr::knit_hooks$set(output = function(x, options) {
  if (is.null(options$tab_cap) == F)
    x
  else
    default_output_hook(x,options)
})
```

# Load per-base scores and annotation

Computed in [methods](./methods.html), tabulated by [mktbl.R](./scripts/mktbl.R)

```{r, message = FALSE}
library(tidyverse)
library(magrittr)
library(pander)
library(stargazer)

scores <- read_tsv("./results/phylop.tsv")
scores %<>% mutate(isUTR = isUTR3 | isUTR5)

# quick visual check of brower vs local phylop
plot(select(scores, browser_phylop_allbr, phylop_allbr))
summary(select(scores, starts_with("phylop")))
```

Convenience functions...

```{r}
# phyloP scores are -log10 p-values: negative for acceleration, positive for conservation
score2p <- function(s){
  10^-(sign(s) * s)
}
```

Plotting functions:

```{r}
theme_set(theme_classic())
plot_histogram <- function(scores_df, test_scores, fill_color = "#3fa9f5", binwidth = 0.1){
  ggplot(scores_df) + geom_histogram(
    aes_string(x = test_scores),
    fill = fill_color,
    color = "grey20",
    binwidth = binwidth
  ) +
  xlab("PhyloP score")
}
```

## All-branches test

```{r}
allbr_hist <- plot_histogram(scores, "phylop_allbr", fill_color = "#68409c") +
  ylab("# of 27way alignment columns")
allbr_hist
allbr_hist + facet_wrap(~isCDS, labeller = label_both)
allbr_hist + facet_wrap(~isIntron, labeller = label_both)
allbr_hist + facet_wrap(~isUTR, labeller = label_both)
allbr_hist + facet_wrap(~isConsEl, labeller = label_both)
allbr_hist + facet_grid(isConsEl~isIntrgn, labeller = label_both)
```

## melanogaster branch test

```{r}
mel_hist <- plot_histogram(scores, "phylop_dmel") +
  ylab(expression(paste("# of ", italic("D. mel"), " Bases")))
mel_hist
mel_hist + facet_wrap(~isCDS, labeller = label_both)
mel_hist + facet_wrap(~isIntron, labeller = label_both)
mel_hist + facet_wrap(~isUTR, labeller = label_both)
mel_hist + facet_grid(isConsEl~isIntrgn, labeller = label_both)
```

## simulans branch test

```{r}
sim_hist <- plot_histogram(scores, "phylop_dsim", fill_color = "#ff1d25") +
  ylab(expression(paste("# of ", italic("D. sim"), " Bases")))
sim_hist
sim_hist + facet_wrap(~isCDS, labeller = label_both)
sim_hist + facet_wrap(~isIntron, labeller = label_both)
sim_hist + facet_wrap(~isUTR, labeller = label_both)
sim_hist + facet_grid(isConsEl~isIntrgn, labeller = label_both)
```

## Final histograms

```{r final-hists-chunk}
mylabs <- as_labeller(
  c(
    "phylop_allbr" = "All~branches",
    "phylop_dmel" = paste("Branch:~", expression(italic("D. mel"))),
    "phylop_dsim" = paste("Branch:~", expression(italic("D. sim")))
  ),
  label_parsed
)

scores_m <- select(scores, -browser_phylop_allbr) %>%
  gather(branch, phylop, contains("phylop"))

allscores <- ggplot(scores_m, aes(x = phylop, fill = branch), color = "grey20") +
  geom_histogram(binwidth = 0.1) +
  facet_wrap(
    ~branch,
    labeller = mylabs
  ) +
  scale_fill_manual(
    guide = FALSE,
    values = c(
      "phylop_allbr" = "#68409c",
      "phylop_dmel" = "#3fa9f5",
      "phylop_dsim" = "#ff1d25"
    )
  ) +
  xlab("PhyloP score") +
  ylab("# of 27way alignment columns")
allscores
ggsave(plot = allscores, "figs/allscores-phylop-scores-hist-3.7k.pdf")
ggsave(plot = allbr_hist + scale_y_continuous(breaks = seq(0, 2000, 500)) + coord_cartesian(ylim = c(0,2200)), "figs/allbr-phylop-scores-hist-3.7k.pdf")
ggsave(plot = mel_hist + scale_y_continuous(breaks = seq(0, 2000, 500)) + coord_cartesian(ylim = c(0,2200)), "figs/dm6-phylop-scores-hist-3.7k.pdf")
ggsave(plot = sim_hist + scale_y_continuous(breaks = seq(0, 2000, 500)) + coord_cartesian(ylim = c(0,2200)), "figs/droSim1-phylop-scores-hist-3.7k.pdf")
```

### counts of accelerated sites

- A score of 2 or -2 corresponds to P-value of 0.01
- Negative := acceleration
- Positive := conservation

```{r}
sig_accel_counts <- scores %>%
  count(
    allbr = phylop_allbr < -2,
    dmel = phylop_dmel < -2,
    dsim = phylop_dsim < -2,
    isConsEl
  )
```

```{r, tab_cap = "Accelerated counts"}
knitr::kable(sig_accel_counts)
```

```{r pander-f-chunk}
pander(ftable(xtabs(n ~ ., sig_accel_counts)), style = "rmarkdown")
```

# Load per-feature tests

```{r per-feature-chunk, message = FALSE}
ftr_fns <- list.files(path="reg_3.7/phylop", pattern = "-phylop27-ftr-\\w+\\.tsv", full.names = TRUE)
ftr_tbl <- lapply(ftr_fns, read_tsv) %>%
  setNames(., nm = sub("\\..*$", "", basename(ftr_fns))) %>%
  bind_rows(.id = "fn")
ftr_tbl %<>%
  separate(fn, c("Branch", "Feature"), "-phylop27-ftr-")
ftr_tbl %>% select(
  Branch, Feature,
  scale, lnlratio, pval,
  null_scale, alt_scale, alt_subscale
)
```

```{r per-feature-p-chunk, tab_cap = "phyloP P-values", message = FALSE}
ftr_tbl %>%
  select(Branch, Feature, pval) %>%
  spread(Branch, pval) %>%
  knitr::kable()
```

<script>
// add bootstrap table styles to pander tables
function bootstrapStylePanderTables() {
  $('tr').parent().parent('table').addClass('table table-condensed');
}
$(document).ready(function () {
  bootstrapStylePanderTables();
});
</script>
