---
title: "Supplemental Methods: Fly mating evolution figures"
author: "Aram Avila-Herrera"
date: "`r date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 4
    code_folding: hide
    highlight: pygments
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = '#>')
sanitizeRSession <- function(s){
  s$platform <- ""
  s$running <- ""
  s$R.version$platform <- ""
  s$R.version$os <- ""
  s$R.version$system <- ""
  s$BLAS <- basename(s$BLAS)
  s$LAPACK <- basename(s$LAPACK)
  s
}
oldSessionInfo <- sessionInfo
sessionInfo <- function(){sanitizeRSession(s = oldSessionInfo())}
```

```{r, message = FALSE}
library(tidyverse)
library(magrittr)
```

Convenience functions:

```{r}
# phyloP scores are -log10 p-values of a likelihood ratio test: negative for acceleration, positive for conservation
score2p <- function(s){
  10^-(sign(s) * s)
}

sem <- function(x){
  sd(x)/sqrt(length(x))
}

kbl_rename <- function(x){
  colnames(x) <- sub("phylop_dmel", "D. mel", colnames(x))
  colnames(x) <- sub("dm6", "D. mel", colnames(x))
  colnames(x) <- sub("phylop_dsim", "D. sim", colnames(x))
  colnames(x) <- sub("droSim1", "D. sim", colnames(x))
  colnames(x) <- sub("phylop_allbr", "All branches", colnames(x))
  colnames(x) <- sub("browser_phylop_allbr", "All branches (browser)", colnames(x))
  return(x)
}

kbl_styl <- function(x, ..., caption){
  knitr::kable(x, caption = caption, format = "html") %>%
  kableExtra::kable_styling(full_width = FALSE, position = "left", ...)
}
```

Plotting functions:

```{r}
theme_set(theme_classic())
plot_histogram <- function(scores_df, test_scores, fill_color = "#3fa9f5", binwidth = 0.1){
  ggplot(scores_df) + geom_histogram(
    aes_string(x = test_scores),
    fill = fill_color,
    color = "grey20",
    binwidth = binwidth
  ) +
  xlab("PhyloP score")
}
```

# Basewise PhyloP scores

## Load basewise scores and annotation

Computed in [methods](./methods.html), tabulated by [mktbl.R](./scripts/mktbl.R)

```{r, message = FALSE}
scores <- read_tsv("./results/phylop.tsv")
scores %<>% mutate(isUTR = isUTR3 | isUTR5)

# quick visual check of brower vs local phylop
plot(select(scores, browser_phylop_allbr, phylop_allbr))
summary(select(scores, contains("allbr")))
```

## Final histograms

(Fig S2DEF: all-branches test scores, subtree test scores for *D. melanogaster* and *D. simulans*)

```{r final-hists-chunk}
mylabs <- as_labeller(
  c(
    "browser_phylop_allbr" = "27~insects",
    "phylop_dmel" = paste(expression(italic("D. melanogaster"))),  # paste("foo~", expression(italic("bar")))
    "phylop_dsim" = paste(expression(italic("D. simulans")))
  ),
  label_parsed
)

scores_m <- select(scores, -phylop_allbr) %>%
  gather(branch, phylop, starts_with("phylop_d"), browser_phylop_allbr)

allscores <- ggplot(scores_m, aes(x = phylop, fill = branch), color = "grey20") +
  geom_histogram(binwidth = 0.1) +
  facet_wrap(
    ~branch,
    labeller = mylabs
  ) +
  scale_fill_manual(
    guide = FALSE,
    values = c(
      "browser_phylop_allbr" = "#68409c",
      "phylop_dmel" = "#3fa9f5",
      "phylop_dsim" = "#dc913c"
    )
  ) +
  xlab("PhyloP score") +
  ylab("# of 27way alignment columns")
allscores
ggsave(plot = allscores, "figs/allscores-phylop-scores-hist-3.7k.pdf")
```

## All-branches test

(alternate Fig S2D)

```{r}
br_allbr_hist <- plot_histogram(scores, "browser_phylop_allbr", fill_color = "#68409c") +
  ylab("# of 27way alignment columns")
br_allbr_hist
ggsave(plot = br_allbr_hist + scale_y_continuous(breaks = seq(0, 2000, 500)) + coord_cartesian(ylim = c(0, 2200)), "figs/browser_allbr-phylop-scores-hist-3.7k.pdf")
ggsave(plot = br_allbr_hist + scale_y_continuous(breaks = seq(0, 200, 50)) + coord_cartesian(ylim = c(0, 210)), "figs/browser_allbr-phylop-scores-hist-3.7k-zoomed.pdf")
```

### By Annotation

#### phastCons Elements

```{r}
scores %>%
  group_by(isConsEl) %>%
  summarize_at(vars(starts_with('browser_phylop_allbr')), funs(mean, sem)) %>%
  kbl_rename() %>%
  kbl_styl(caption = "PhyloP scores")
```

```{r}
br_allbr_hist + facet_wrap(~isConsEl, labeller = label_both)
```

#### Coding sequences

```{r}
scores %>%
  group_by(isCDS) %>%
  summarize_at(vars(starts_with('browser_phylop_allbr')), funs(mean, sem)) %>%
  kbl_rename() %>%
  kbl_styl(caption = "PhyloP scores")
```

```{r}
br_allbr_hist + facet_wrap(~isCDS, labeller = label_both)
```

#### UTRs

```{r}
scores %>%
  group_by(isUTR) %>%
  summarize_at(vars(starts_with('browser_phylop_allbr')), funs(mean, sem)) %>%
  kbl_rename() %>%
  kbl_styl(caption = "PhyloP scores")
```

```{r}
br_allbr_hist + facet_wrap(~isUTR, labeller = label_both)
```

#### Introns

```{r}
scores %>%
  group_by(isIntron) %>%
  summarize_at(vars(starts_with('browser_phylop_allbr')), funs(mean, sem)) %>%
  kbl_rename() %>%
  kbl_styl(caption = "PhyloP scores")
```

```{r}
br_allbr_hist + facet_wrap(~isIntron, labeller = label_both)
```

#### Intergenic

```{r}
scores %>%
  group_by(isIntergenic) %>%
  summarize_at(vars(starts_with('browser_phylop_allbr')), funs(mean, sem)) %>%
  kbl_rename() %>%
  kbl_styl(caption = "PhyloP scores")
```

```{r}
br_allbr_hist + facet_wrap(~isIntergenic, labeller = label_both)
```
```

## Subtree tests

### Melanogaster

(alternate Fig S2E)

```{r}
mel_hist <- plot_histogram(scores, "phylop_dmel") +
  ylab(expression(paste("# of ", italic("D. mel"), " Bases")))
mel_hist
ggsave(plot = mel_hist + scale_y_continuous(breaks = seq(0, 2000, 500)) + coord_cartesian(ylim = c(0, 2200)), "figs/dm6-phylop-scores-hist-3.7k.pdf")
```

### Simulans

(alternate Fig S2F)

```{r}
sim_hist <- plot_histogram(scores, "phylop_dsim", fill_color = "#dc913c") +
  ylab(expression(paste("# of ", italic("D. sim"), " Bases")))
sim_hist
ggsave(plot = sim_hist + scale_y_continuous(breaks = seq(0, 2000, 500)) + coord_cartesian(ylim = c(0, 2200)), "figs/droSim1-phylop-scores-hist-3.7k.pdf")
```

### By Annotation

```{r}
subtreescores <- ggplot(
    filter(scores_m, grepl("^phylop_d", branch)),
    aes(x = phylop, fill = branch), color = "grey20"
  ) +
  geom_histogram(binwidth = 0.1) +
  facet_wrap(
    ~branch,
    labeller = mylabs
  ) +
  scale_fill_manual(
    guide = FALSE,
    values = c(
      #"browser_phylop_allbr" = "#68409c",
      "phylop_dmel" = "#3fa9f5",
      "phylop_dsim" = "#dc913c"
    )
  ) +
  xlab("PhyloP score") +
  ylab("# of 27way alignment columns")
```

#### phastCons Elements

```{r}
scores %>%
  group_by(isConsEl) %>%
  summarize_at(vars(starts_with('phylop_d')), funs(mean, sem)) %>%
  kbl_rename() %>%
  kbl_styl(caption = "PhyloP scores")
```

```{r}
subtreescores + facet_grid(isConsEl~branch, labeller = label_both)
```

#### Coding sequences

```{r}
scores %>%
  group_by(isCDS) %>%
  summarize_at(vars(starts_with('phylop_d')), funs(mean, sem)) %>%
  kbl_rename() %>%
  kbl_styl(caption = "PhyloP scores")
```

```{r}
subtreescores + facet_grid(isCDS~branch, labeller = label_both)
```

#### UTRs

```{r}
scores %>%
  group_by(isUTR) %>%
  summarize_at(vars(starts_with('phylop_d')), funs(mean, sem)) %>%
  kbl_rename() %>%
  kbl_styl(caption = "PhyloP scores")
```

```{r}
subtreescores + facet_grid(isUTR~branch, labeller = label_both)
```

#### Introns

```{r}
scores %>%
  group_by(isIntron) %>%
  summarize_at(vars(starts_with('phylop_d')), funs(mean, sem)) %>%
  kbl_rename() %>%
  kbl_styl(caption = "PhyloP scores")
```

```{r}
subtreescores + facet_grid(isIntron~branch, labeller = label_both)
```

#### Intergenic

```{r}
scores %>%
  group_by(isIntergenic) %>%
  summarize_at(vars(starts_with('phylop_d')), funs(mean, sem)) %>%
  kbl_rename() %>%
  kbl_styl(caption = "PhyloP scores")
```

```{r}
subtreescores + facet_grid(isIntergenic~branch, labeller = label_both)
```


### Counts of accelerated sites

- A score of 2 or -2 corresponds to an unadjusted P-value of 0.01 for that base
- Negative := acceleration
- Positive := conservation

```{r}
sig_accel_counts <- scores %>%
  count(
    "All branches" = phylop_allbr < -2,
    "D. mel" = phylop_dmel < -2,
    "D. sim" = phylop_dsim < -2,
    "Is Cons El" = isConsEl
  )
```

```{r}
kbl_styl(sig_accel_counts, caption = "Counts of accelerated sites")
```

```{r pander-f-chunk}
pander::pander(ftable(xtabs(n ~ ., sig_accel_counts)), style = "rmarkdown")
```

# Feature-type PhyloP tests

## Load per-feature-type results

```{r per-feature-chunk, message = FALSE}
ftr_fns <- list.files(
  path="reg_3.7/phylop",
  pattern = "-phylop27-ftr-\\w+\\.tsv",
  full.names = TRUE
)
ftr_tbl <- lapply(ftr_fns, read_tsv) %>%
  setNames(., nm = sub("\\..*$", "", basename(ftr_fns))) %>%
  bind_rows(.id = "fn")
ftr_tbl %<>%
  separate(fn, c("Branch", "FeatureType"), "-phylop27-ftr-")
ftr_tbl %>% select(
  Branch, FeatureType,
  scale, lnlratio, pval,
  null_scale, alt_scale, alt_subscale
)
```

```{r per-feature-p-chunk, message = FALSE}
ftr_tbl %>%
  select(Branch, FeatureType, pval) %>%
  spread(Branch, pval) %>%
  rename(
    `All branches` = allbr,
    `D. mel` = dm6,
    `D. sim` = droSim1
  ) %>%
  kbl_styl(caption = "phyloP P-values" )
```

# Whole-region all-branches conservation test

```{r}
read_tsv("reg_3.7/phylop/allbr-phylop27-wholeregion-con.tsv") %>%
  kbl_styl(caption = "PhyloP test for conservation")
```


# R session info

```{r}
s <- sessionInfo()
s
```

# R Citations

```{r}
lapply(
  c(
    s$basePkgs,
    names(s$otherPkgs),
    names(s$loadedOnly)
  ),
  citation
)
```

<script>
// add bootstrap table styles to pander tables
function bootstrapStylePanderTables() {
  $('tr').parent().parent('table').addClass('table table-condensed');
}
$(document).ready(function () {
  bootstrapStylePanderTables();
});
</script>
