---
title: "Supplemental Methods: Fly mating evolution"
author: "Aram Avila-Herrera"
date: "`r date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 4
    code_folding: hide
    highlight: pygments
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = '#>')
sanitizeRSession <- function(s){
  s$platform <- ""
  s$running <- ""
  s$R.version$platform <- ""
  s$R.version$os <- ""
  s$R.version$system <- ""
  s$BLAS <- basename(s$BLAS)
  s$LAPACK <- basename(s$LAPACK)
  s
}
oldSessionInfo <- sessionInfo
sessionInfo <- function(){sanitizeRSession(s = oldSessionInfo())}
```

# gr32a regulatory region

Referenced in the final manuscript as "3.8 kbp regulatory region" (and in
working drafts as "3.7 kbp"); it contains a gene (CG6201) transcribed in the
opposite direction of gr32a. It is 3,798 bp long.  We had previously analyzed a
661 bp sub-region immediately upstream gr32a.

## Coordinates

#### [dm6: `chr2L:11,110,412-11,114,209`](https://genome.ucsc.edu/cgi-bin/hgTracks?db=dm6&position=chr2L%3A11110412-11114209)

Browser coordinates obtained by BLAT search of *melanogaster* sequence
`>Gr32[mel]` in file: `Gr32aReg.rtf` (copied to `gr32a_3.7kb_upstrm.fa`)
against the *dm6* melanogaster genome assembly.

**Please note: UCSC browser coordinates start at 1 and include the end coordinate in the specified interval, i.e.
"1-offset closed". MAF and BED formats start counting with 0 (0-offset). BED
format does not consider the end coordinate to be in the interval---it is
"right-open"**

Intron/CDS/UTR/intergenic regions were obtained from the UCSC Table Browser in
BED format. "Intergenic" is the 3.8kbp region minus "whole gene" features. I
also downloaded the phastCons conserved elements.

```{bash}
bedtools subtract -a reg_3.7/coords/reg_3.7.bed -b reg_3.7/coords/wholegene.bed > reg_3.7/coords/intergenic.bed
```

### List of subregions:

Intron, cds exon, 5' UTR, 3' UTR, intergenic, 661 bp ("gr32-upstream-adjacent")

```{bash}
head reg_3.7/coords/{cds_exons,introns,utr3,utr5,intergenic,661}.bed
```

Conserved elements

```{bash}
gzcat reg_3.7/coords/dm6-allbr-consel27.bed.gz | head
```

## PhyloP

UCSC Browser track shows conservation/acceleration scores for the coordinates
above (`dm6-allbr-phylop27.wig.gz` downloaded from the Table Browser). PhyloP
was run as follows:

- Performs an all branch test (no `--subtree`)
- Score for each base (`--wig-scores`)
- Scores are -log p-values for likelihood ratio tests (`--method LRT`)
- Conservation and acceleration scores reported (`--mode CONACC`)
- Uses the phastCons neutral model, but adjusts the background frequencies to
  make the model reversible.

### Data

Alignment

: 27-way alignment for 23 *Drosophila* species, housefly, mosquito, honeybee,
  and beetle, downloaded from the Table Browser for the coordinates above
  (`reg_3.7/alignments/dm6-allbr-27way.maf`).
  <http://hgdownload.cse.ucsc.edu/goldenPath/dm6/multiz27way/>

Neutral model

: Phylogenetic model for neutral evolution from 4-fold degenerate sites in
  alignment. `dm6.phyloP27way.mod` from
  <http://hgdownload.cse.ucsc.edu/goldenPath/dm6/phyloP27way/>

See also: <https://genome.ucsc.edu/cgi-bin/hgTrackUi?g=cons27way&db=dm6>

### Verify methods against browser track

Run `phylop` as above to check that we can reproduce the browser track
(all-branches, basewise test).

```{bash}
phylop --wig-scores --method LRT --mode CONACC dm6.phyloP27way.mod reg_3.7/alignments/dm6-allbr-27way.maf > reg_3.7/phylop/allbr-phylop27.wig
```

Approximately the same, but not exact... (possibly due to optimization seed and
rounding)

```{bash}
head reg_3.7/phylop/allbr-phylop27.wig
```

```{bash}
gzcat reg_3.7/tracks/dm6-allbr-phylop27.wig.gz | head -n 20  # browser track
```


### Subtree basewise tests

We are interested in accelerated positions in the *melanogaster* and *simulans*
branches. Especially, if they are in regions conserved across fruit flies,
suggesting that something interesting may be occuring in a functional
regulatory region.

- Re-run `phylop` with `--subtree dm6` and `--subtree droSim1`

```{bash}
phylop --subtree dm6 --wig-scores --method LRT --mode CONACC dm6.phyloP27way.mod reg_3.7/alignments/dm6-allbr-27way.maf > reg_3.7/phylop/dm6-phylop27.wig
phylop --subtree droSim1 --wig-scores --method LRT --mode CONACC dm6.phyloP27way.mod reg_3.7/alignments/dm6-allbr-27way.maf > reg_3.7/phylop/droSim1-phylop27.wig
```

- Make table ([**`phylop.tsv`**](./phylop.tsv), [mktbl.R](./scripts/mktbl.R)) of
  phyloP scores and regions:
    - pos (browser coordinate)
    - isConsEl
    - phylop_allbr
    - phylop_dmel
    - phylop_dsim
    - browser_phylop_allbr
    - is661
    - isUTR3
    - isUTR5
    - isCDS
    - isIntron
    - isIntergenic

```{bash}
scripts/mktbl.R reg_3.7/ 2> results/mktbl.stderr > results/mktbl.stdout
head results/phylop.tsv
```

- Extra/TODO: Re-call phastCons elements for non-*melanogaster* and non-*simulans* alignments.
    - isConsEl_nomel
    - isConsEl_nosim

- Histogram of p-values for mel, sim
  **Figures:** See [./figs.html](./figs.html)

### Feature tests

- Re-run `phylop` with `--features` (one score for each exon, intron, utr,
  intergenic region)

```{bash}
# Be explicit about trimming and sorting features
cat reg_3.7/coords/{cds_exons,introns,utr3,utr5,intergenic}.bed | \
  cut -f1-4 | \
  bedtools intersect -a - -b reg_3.7/coords/reg_3.7.bed | \
  sort -u -k1,3 | \
  sort-bed - > reg_3.7/coords/features.bed
head reg_3.7/coords/features.bed
```

#### All-branches

```{bash}
phylop --features reg_3.7/coords/features.bed --method LRT --mode CONACC dm6.phyloP27way.mod reg_3.7/alignments/dm6-allbr-27way.maf > reg_3.7/phylop/allbr-phylop27-features.tsv
```

#### Subtree

```{bash}
phylop --features reg_3.7/coords/features.bed --subtree dm6 --method LRT --mode CONACC dm6.phyloP27way.mod reg_3.7/alignments/dm6-allbr-27way.maf > reg_3.7/phylop/dm6-phylop27-features.tsv

phylop --features reg_3.7/coords/features.bed --subtree droSim1 --method LRT --mode CONACC dm6.phyloP27way.mod reg_3.7/alignments/dm6-allbr-27way.maf > reg_3.7/phylop/droSim1-phylop27-features.tsv
```

### Feature-type tests

- Get one score/p-value per combined UTRs, combined CDSs, combined introns,
  combined intergenic regions

```{bash}
mmaf=reg_3.7/alignments/dm6-allbr-27way.maf
cat reg_3.7/coords/utr[35].bed > reg_3.7/coords/utr.bed
for ftr in cds_exons utr introns intergenic; do
  fmaf=reg_3.7/alignments/dm6-allbr-27way-$ftr.maf
  bed=reg_3.7/coords/${ftr}.bed
  maf_parse --features ${bed} ${mmaf} > ${fmaf}

  # all-branches
  phylop --features reg_3.7/coords/reg_3.7.bed --method LRT --mode CONACC -i MAF dm6.phyloP27way.mod ${fmaf} > reg_3.7/phylop/allbr-phylop27-ftr-${ftr}.tsv

  # subtree
  phylop --subtree dm6 --features reg_3.7/coords/reg_3.7.bed --method LRT --mode CONACC -i MAF dm6.phyloP27way.mod ${fmaf} > reg_3.7/phylop/dm6-phylop27-ftr-${ftr}.tsv

  phylop --subtree droSim1 --features reg_3.7/coords/reg_3.7.bed --method LRT --mode CONACC -i MAF dm6.phyloP27way.mod ${fmaf} > reg_3.7/phylop/droSim1-phylop27-ftr-${ftr}.tsv
done
```

### Whole-region tests

#### All-branches conservation

```{bash}
phylop --features reg_3.7/coords/reg_3.7.bed --method LRT --mode CON dm6.phyloP27way.mod reg_3.7/alignments/dm6-allbr-27way.maf > reg_3.7/phylop/allbr-phylop27-wholeregion-con.tsv
```

---

# Other notes

## adjacent region (661bp)

Our previous analysis focused on the 661bp region immediately upstream gr32a
(technically, includes a bit of the 5' UTR).

`gr32a_661_upstrm.fa`

### "New"

Same methodology as "Old", just uses newer assemblies, alignments, and
phylogenetic models. We also now look at the whole 3.8kb region and compare
exonic vs non-exonic regions.

- `dm6-phylop27.wig`
- `droSim1-phylop27.wig`

### "Old"

- dm3-conacc-allbr track (661bp):
    - chr2L:11,113,549:11,114,209
    - species: 12 flies, mosquito, honeybee, beetle
    - test for acceleration or conservation, base-by-base
    - score is a -log10 p-value of a likelihood ratio test between "fixed" vs
      "free" tree scaling parameter evolution models.
      "negative" scores correspond to acceleration, positive to conservation.
      e.g., 1.7 = -log10(p-value) => 0.02 = p-value;
- dmel: same region, but test is only on melanogaster branch (dm3)
- droSim1: same region, but test is only on simulans branch (droSim1)

# Versions and Citations

## bedtools

Quinlan AR and Hall IM, 2010. BEDTools: a flexible suite of utilities for
comparing genomic features. Bioinformatics. 26, 6, pp. 841–842.

```{bash}
bedtools --version
```

## bedops

```{bash}
bedops --version
```

## phast package

```{bash}
cat ~/src/phast/version
phylop -h | tail -n 12
```

## R session info

```{r}
s <- sessionInfo()
s
```

## R Citations

```{r}
lapply(
  c(
    s$basePkgs,
    names(s$otherPkgs),
    names(s$loadedOnly)
  ),
  citation
)
```
